\documentclass[11pt]{article}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage[margin=1in]{geometry}
\usepackage{float}
\usepackage{helvet}
\renewcommand{\familydefault}{\sfdefault}
\usepackage[font=sf,labelfont=bf,skip=5pt]{caption}
\usepackage{hyperref}
\hypersetup{colorlinks=true,linkcolor=blue,urlcolor=blue,citecolor=blue}
\usepackage{setspace}
\onehalfspacing
\usepackage{authblk}
\usepackage{longtable}
\usepackage{pdflscape}

\title{Learning Normal Representations for Blood Biomarkers}

% Add authors here
% \author[1]{Author One}
% \author[2]{Author Two}
% \affil[1]{Department, Institution}
% \affil[2]{Department, Institution}

\date{}

\begin{document}

\maketitle

% ═══════════════════════════════════════════════════════════════════
% ABSTRACT
% ═══════════════════════════════════════════════════════════════════

\begin{abstract}
Clinical laboratories often interpret blood tests against fixed population reference intervals that assume a single normal range for all individuals. However, most biomarkers fluctuate around narrow personal setpoints, so population-normal values can mask clinically meaningful deviation from an individual's baseline, risking delayed disease detection. Purely personalized intervals derived from individual histories address this but can overfit to sparse data, inflating false-positive rates and unnecessary follow-up. Here we introduce NORMA, a sequence model trained on 3.4M longitudinal laboratory sequences that reframes lab interpretation with conditional prediction, asking: given a patient's history, what lab values would be expected if the patient is in a ``healthy'' state? We validate NORMA-derived reference intervals (NORMA-RI) across two cohorts: a national observational cohort in Clalit Health Services (3.5M adults, 1.9B measurements) with over two decades and an acute care cohort in the eICU Collaborative Research Database (98,432 ICU patients, over 20M measurements). Across both settings, NORMA-RI detected more abnormalities than population intervals (16.1\% versus 8.4\% in CHS; 56.0\% versus 50.3\% in eICU) while flagging far fewer than personalized intervals (28.4\% in CHS; 68.1\% in eICU). Among measurements missed by population intervals, NORMA-RI improved prediction accuracy over PerRI for eICU outcomes (in-hospital mortality, AKI, sepsis, prolonged length of stay) and for delayed outcomes in CHS (all-cause mortality, type 2 diabetes, chronic kidney disease) over a 10-year horizon. By anchoring reference ranges to both population-level distributions and individual longitudinal trajectories, NORMA delivers personalization without overflagging, enhancing detection of meaningful change at scale. To promote transparency, we publicly release the model, code, and an interactive user interface for accessible, individualized laboratory interpretation.
\end{abstract}

\clearpage

% ═══════════════════════════════════════════════════════════════════
% INTRODUCTION
% ═══════════════════════════════════════════════════════════════════

\section*{Introduction}

Laboratory testing is among the most frequently performed medical procedures, with more than 14 billion tests ordered annually in the United States alone\textsuperscript{1}. Yet interpretation has changed remarkably little since the 1960s; values are classified as ``low,'' ``normal,'' or ``high'' against fixed population reference intervals (Pop-RI) that represent the central 95\% of measurements from ostensibly healthy individuals\textsuperscript{1--3}. Prior studies have recognized that many routine biomarkers fluctuate around narrow individual setpoints and that within-person change carries prognostic information that population intervals miss\textsuperscript{4--8}. Despite this, clinical practice still relies on universal thresholds, partly because of the ease of a one-size-fits-all approach, and because operationalizing individualized interpretation requires modeling patient-specific trajectories at scale.

Efforts to move beyond universal thresholds have taken two directions. The first refines reference intervals for predefined subgroups, for example, sex-specific hematologic ranges or adjusted HbA1c thresholds for patients with hemoglobin variants\textsuperscript{9--11}. While clinically useful, such definitions of subgroups are often defined coarsely or arbitrarily, may encode existing biases, and cannot capture the full spectrum of individual variation embedded in longitudinal trajectories. The second derives personalized reference intervals (Per-RI) directly from an individual's laboratory history\textsuperscript{8}. Foy et al.\ showed that deviations from individually derived hematologic setpoints associate more strongly with mortality than deviations from population thresholds, underscoring the prognostic value of within-person change\textsuperscript{8}. However, purely individualized approaches that ignore population-level trends may incorporate unrecognized chronic disease into the estimated baseline and risk overfitting to sparse personal data, producing overly narrow intervals that flag benign variation and inflate false-positive rates\textsuperscript{8,12,13}.

Here, we introduce NORMA (Normal Outcome Range Modeling with Attention), which reframes laboratory interpretation as a conditional prediction problem: given a patient's longitudinal history, what value would be expected at the next measurement if the patient remains in a healthy state? Trained on more than 3.4 million longitudinal laboratory sequences, NORMA generates individualized reference intervals for 30 common blood analytes by learning to anchor patient-specific trajectories to population-level distributions (Figure~1A--C). We externally validate NORMA in two distinct settings: an outpatient cohort of 3.5 million adults in Clalit Health Services with 1.7 billion test results over two decades, and an acute care cohort of over a hundred thousand ICU admissions in the eICU Collaborative Research Database with 25 million laboratory measurements. By learning temporal patterns from longitudinal sequences and anchoring individualized predictions to population-level distributions, NORMA resolves the tradeoff between the false reassurance of PopRI and the overcalling of PerRI, regardless of clinical setting.


% ═══════════════════════════════════════════════════════════════════
% RESULTS
% ═══════════════════════════════════════════════════════════════════

\section*{Results}

\subsection*{Study populations}

We assessed reference intervals across four common clinical panels: complete blood count, comprehensive metabolic panel, lipid profile, and HbA1c. Development cohorts (MIMIC-IV and EHRSHOT) spanned inpatient and outpatient settings with comparable laboratory value distributions (Supplementary Table~2). External validation comprised two cohorts that differ in temporal scale and clinical setting. After filtering for a minimum of five measurements per test, the Clalit Health Services (CHS) cohort included 903,022 adults and 16,301,161 sequences between 2000 and 2024, with clinically stable trajectories spanning a median of more than four years (Figure~1B, 2A). The eICU Collaborative Research Database cohort included 98,432 ICU patients across 208 U.S.\ hospitals contributing 1,135,814 laboratory trajectories spanning a median of $\sim$8 days but densely sampled (median 8--35 measurements per patient for high-frequency analytes). Laboratory distributions in CHS largely overlapped conventional population reference intervals (Supplementary Table~2), whereas the eICU cohort exhibited shifts consistent with critical illness physiology, including lower albumin and higher glucose variability.

\subsection*{Lab values are individualized}

In both validation cohorts (CHS and eICU), within-person variability was substantially smaller than between-person variability for most analytes, confirming that laboratory values fluctuate around narrow individual setpoints rather than spanning the full population range (Figure~2B--D). In CHS, intra-individual coefficients of variation were consistently lower than inter-individual coefficients, with analytes such as alkaline phosphatase, creatinine, and LDL cholesterol showing the most stable setpoints (intra-/inter-CV ratio $<$0.40). In eICU, indices of individuality were generally higher, reflecting the physiological of patients with critical illness, yet hematologic indices such as MCH, RDW, and MCV maintained low individuality (intra-/inter-CV ratio $<$0.25) even in the ICU, suggesting that personalized intervals remain informative for these analytes regardless of clinical setting. In contrast, potassium and glucose exhibited higher intra-patient variability in the ICU, indicating that for these analytes, within-person variability approached the population spread, consistent with rapid physiological fluctuations during critical illness.

This within-person stability carried prognostic signal: both where a measurement falls within the population distribution and how far it departs from a patient's personal setpoint were related to mortality risk. In CHS and eICU, both the absolute level of a biomarker and the magnitude of deviation from baseline were monotonically associated with risk, with several analytes, including WBC, glucose, AST, and HbA1c, exhibiting U-shaped mortality curves (Supplementary Figure~1, Figure~2C).

\subsection*{NORMA generates calibrated individualized intervals}

NORMA is a conditional, autoregressive decoder-only model which uses longitudinal laboratory trajectories to predict the distribution of a patient's next measurement (Figure~3A; Methods). We trained NORMA on 3.4 million longitudinal sequences spanning 30 analytes from MIMIC-IV and EHRSHOT (Figure~3A; Supplementary Methods; Supplementary Table~2). We trained NORMA under two output parameterizations, Gaussian and quantile, which share the same architecture but differ in distributional assumptions (see Methods); results were consistent across both configurations (Supplementary Table~4). From this predicted distribution of a patient's next measurement, we use the 95\% percentile to define a reference interval, NORMA\textsubscript{RI}.

Accurate next-step prediction is a prerequisite for meaningful reference intervals. On the held-out test set, NORMA achieved a mean absolute percentage error of 9.4 $\pm$ 0.02 across analytes (versus 12.3 $\pm$ 0.5 for the best baseline) and the highest coefficient of determination (R$^2$ = 0.74 $\pm$ 0.12 versus 0.58 $\pm$ 0.20 for ARIMA) by using the distribution's midpoint (Figure~3B--C). Forecasting accuracy varied among analytes, with MPV showing the weakest performance, likely due to it having the smallest training sample size among the 30 analytes.

Simulations using baseline trajectories demonstrate that NORMA's prediction intervals responded appropriately to the factors that should govern uncertainty (Figure~3D). Increasing within-person variability widened intervals (from 0.1\% to 93\% of the population reference range as within-person SD increased from 0\% to 30\%), additional prior measurements narrowed them (median 111\% reduction from 2 to 300 measurements, with 80\% of the gain in the first 30), and longer forecast horizons produced monotonic widening (median 49\% increase from days to 10 years). Intervals also shifted with patient age and sex in directions consistent with known physiological differences (Supplementary Figure~8).

When applied to the validation cohorts, NORMA-RI consistently occupied an intermediate position between population and purely personalized intervals. In CHS, among individuals with no abnormalities flagged by population reference ranges, NORMA-RI flagged an average of 2 additional tests per person compared to 3 for PerRI, increasing the abnormality rate from 8.4\% (PopRI) to 16.1\% (NORMA-RI) versus 28.4\% (PerRI; Figure~4A). In eICU, where baseline abnormality rates were substantially higher reflecting acute illness, the same ordering held: 50.3\% (PopRI), 56.0\% (NORMA-RI), and 68.1\% (PerRI), demonstrating that NORMA's intermediate behavior is robust to both the long, sparse trajectories of a national health system and the short, dense trajectories of acute care.

\subsection*{NORMA-RI improve clinical prediction}

In eICU, when NORMA flagged abnormalities earlier than PopRI, the median lead time was 25.2 hours (IQR: 22.3--50.7 hours). For glucose, the most frequently sampled analyte in critical care, NORMA flagged abnormalities a median of 9.6 hours earlier. For analytes measured less frequently, such as albumin (47.8 hours) and total bilirubin (46.9 hours), the lead time was correspondingly longer. In CHS, where measurements are spaced months to years apart, lead times were on the order of X (X).

We evaluated whether the additional abnormalities detected by NORMA-RI, beyond those already flagged by population intervals, carried clinical meaning. In CHS, we assessed prediction of delayed outcomes (all-cause mortality, type 2 diabetes, and chronic kidney disease) over a 10-year horizon. In eICU, we assessed prediction of immediate outcomes (in-hospital mortality, acute kidney injury, sepsis, and prolonged ICU length of stay $>$7 days). In both cohorts, we restricted evaluation to measurements classified as normal by population intervals, so that all reported gains reflect NORMA-RI's ability to detect signal that standard clinical practice misses.

Across both settings, PerRI achieved higher sensitivity than NORMA-RI, consistent with its tendency to flag more results as abnormal. However, NORMA-RI achieved higher specificity and precision, the metrics that matter when the clinical cost of a false alarm is high. In CHS, NORMA-RI improved mean specificity and precision relative to PerRI by +11.8\% and +2.7\%, respectively (Figure~4C), with the largest precision gains for albumin and mortality (+12.6\%), glucose and type 2 diabetes (+10.8\%), and creatinine and chronic kidney disease (+6.5\%). In eICU, NORMA-RI achieved mean specificity of 0.83 versus 0.52 for PerRI across outcomes, with higher positive predictive value for mortality (0.125 versus 0.106) and prolonged length of stay (0.273 versus 0.261). NORMA-RI improved precision for the majority of analyte-outcome pairs, with gains most pronounced for analytes with stable individual setpoints; for analytes with high within-person variability such as MPV, the advantage over PerRI was attenuated.

In time-to-event analyses, NORMA-RI outperformed both PerRI and PopRI in time-to-event prediction in eICU and CHS (Figure~4D--5D). These results held across temporal scales spanning days to decades, clinical contexts ranging from national observational care to acute ICU admissions, and outcome horizons from in-hospital events to 10-year disease incidence.


% ═══════════════════════════════════════════════════════════════════
% DISCUSSION
% ═══════════════════════════════════════════════════════════════════

\section*{Discussion}

Routine laboratory testing is becoming more frequent, more accessible, and more central to preventive medical care\textsuperscript{24,25}. Beyond clinician-ordered diagnostics, patients now undergo repeated panel-based screening through health systems and direct-to-consumer services. In parallel, interpretation is slowly shifting from static, population-based cutoffs toward ``personalized'' views that emphasize trajectories and individual baselines\textsuperscript{6,7,17,22}.

Our results illustrate both the promise and the perils of individualized interpretation across two fundamentally different clinical settings. In CHS, a population of nearly one million adults, we found that many routine analytes are substantially more stable within individuals than across individuals, and that deviations from a person's usual range are associated with higher mortality risk\textsuperscript{8}. At the same time, empirical PerRI built solely from an individual's history labeled far more results as abnormal than population-based intervals, which naturally translates to lower positive predictive value and higher false-positive rates for major outcomes.

In the eICU cohort, we extended this evaluation to acute care, where laboratory trajectories are short, densely sampled, and influenced by rapid physiological changes. Despite these challenges, NORMA maintained its core behavior: generating intervals that fall between population and purely personalized thresholds. The ICU setting amplified baseline abnormality rates substantially (50\% vs.\ 8\% under PopRI), yet NORMA still discriminated between clinically meaningful deviations and expected variation. Among measurements that population intervals called normal, NORMA flagged 11.4\% as abnormal, less than half the 29.5\% flagged by PerRI, and among those NORMA uniquely flagged, one in ten patients died in hospital and one in six developed sepsis. Even with baseline trajectories spanning a median of only $\sim$8 days compared to $>$4 years in CHS, NORMA produced calibrated intervals that identified patients at elevated risk of AKI, sepsis, and mortality. This gain in specificity comes at a cost in sensitivity; NORMA detected approximately 20\% of true events compared to 50\% for PerRI. However, this trade-off is inherent as PerRI narrows the set of flagged values.

By conditioning on a ``normal'' health state, the model learns what a patient's next value should be if they remain physiologically stable. Unlike PerRI, which derives its baseline entirely from an individual's history and may therefore absorb chronic disease or acute perturbation into the estimated setpoint, NORMA anchors each patient's trajectory to population-level expectations for healthy individuals. This prevents the baseline contamination that leads PerRI to normalize pathology. At the same time, NORMA's time-aware architecture handles the irregular sampling and temporal clustering inherent to clinical data, a property that proved particularly important in the eICU setting. The consistency of these results across the two validation cohorts is notable: CHS and eICU used different NORMA configurations (Gaussian vs.\ quantile output parameterization, different temporal encodings and age representations; Supplementary Table~3), yet the core finding, that NORMA occupies an intermediate position with higher precision, held across both.

NORMA has several practical advantages over empirical PerRI and Bayesian approaches suggested in prior literature\textsuperscript{22,27}. It does not require explicit pre-filtering of trajectories for stability, it handles irregular measurement spacing natively, and its inputs are limited to data already present in most electronic health records. In a clinical deployment, NORMA could provide ``personalized'' assessment for patients even without sufficient measurement history. The sensitivity analysis showed that the reference range starts wide and narrows with additional measurements, suggesting that patients with even a modest laboratory history can benefit. Unlike PerRI, having too many measurements does not cause over-narrowing of the reference interval, because NORMA's interval width stabilizes after approximately 30 measurements.

There are additional limitations. NORMA was trained and evaluated on 30 common blood tests and conditioned only on age, sex, and laboratory trajectories; it does not currently incorporate comorbidities, medications, or other clinical context, which could further refine expected trajectories and abnormality thresholds. Furthermore, NORMA's performance was not uniform across analytes. MPV showed the lowest performance, consistent with its having the smallest training sample size and weakest forecasting accuracy among the 30 analytes. We modeled predictive distributions as Gaussian in CHS, which may not fully capture skewed or heavy-tailed analytes; however, we implemented quantile regression in eICU and results were consistent across both parameterizations. The CHS analysis was conducted within a single national health system, and the eICU cohort, while geographically diverse, represents only the ICU segment of hospital care; performance may differ in emergency departments, primary care clinics, or populations with different demographics and laboratory utilization patterns. Finally, this study evaluates retrospective associations between abnormality flags and clinical outcomes; prospective studies are needed to test whether interpretation changes clinician behavior, enables earlier disease detection, or reduces unnecessary downstream testing.

Individualized laboratory interpretation can be made more precise and more clinically useful by combining patient-specific trajectories with population-level expectations and uncertainty-aware prediction. By validating across longitudinal outpatient and acute settings, outcome horizons, and patient populations, we show that this framework generalizes beyond a single clinical context. NORMA offers a path toward integrating personalized laboratory interpretation into routine practice, balancing the sensitivity of personalized intervals with the specificity of population-anchored prediction.


% ═══════════════════════════════════════════════════════════════════
% METHODS
% ═══════════════════════════════════════════════════════════════════

\section*{Methods}

\subsection*{Laboratory Measurements}

We analyzed 30 laboratory analytes across four datasets organized in two stages: MIMIC-IV and EHRSHOT for model development (training, validation, and held-out testing), and two independent external validation cohorts, Clalit Health Services (CHS; outpatient, long-horizon) and the eICU Collaborative Research Database (eICU-CRD; critical care, short-horizon). No development data were used in external validation. We mapped all laboratory results to Logical Observation Identifiers Names and Codes (LOINC). We retained only positive values, standardized units to canonical formats, and excluded analyte-specific outliers exceeding $\pm$3 standard deviations from the analyte median. When duplicate results shared identical timestamps, we averaged values within each analyte. We grouped analytes into four standard clinical panels: Complete blood count (hematocrit, hemoglobin, mean corpuscular hemoglobin, mean corpuscular hemoglobin concentration, mean platelet volume, platelet count, red blood cell count, red cell distribution width, white blood cell count); Comprehensive metabolic panel (sodium, potassium, chloride, calcium, bicarbonate, blood urea nitrogen, creatinine, glucose, aspartate aminotransferase, alanine aminotransferase, alkaline phosphatase, total bilirubin, albumin); Lipid panel (total cholesterol, low-density lipoprotein cholesterol, high-density lipoprotein cholesterol, triglycerides); Glycemic control (glycated hemoglobin).

\subsection*{Reference Interval Frameworks}

\textbf{Population Reference Intervals.} We defined PopRI as the central 95\% of values from ostensibly healthy adults without chronic disease using the American Board of Internal Medicine Laboratory Test Reference Ranges (January 2025). Supplementary Table~1 provides sex-specific intervals, LOINC codes, and canonical units for all analytes.

\textbf{Personalized Reference Intervals.} We estimated PerRI following Foy et al.\ For each analyte--patient pair, we fit Gaussian mixture models with one to three components to baseline measurements and selected model order using the Akaike information criterion. We designated the component containing the majority of observations as the individual's physiological setpoint and defined the PerRI as the component mean $\pm$ 2 standard deviations. Eligibility criteria for baseline measurements are described separately for each validation cohort.

\subsection*{NORMA (Normal Outcome Range Modeling with Attention)}

NORMA is a conditional, decoder-only Transformer that models the distribution of a patient's next laboratory value given their longitudinal measurement history and a query specifying a future health state and prediction horizon. We trained the model on 3.4 million longitudinal sequences from MIMIC-IV and EHRSHOT, requiring at least three repeated measurements per patient without constraints on sampling intervals or clinical setting. Data were split at the patient level into training (1,982,150 sequences; 70\%), validation (283,165; 10\%), and test (566,329; 20\%) sets to prevent leakage across partitions. Each sequence was tokenized, padded, and masked to handle variable-length histories. A weighted sampling scheme was used during training to balance representation across 30 laboratory analytes.

\textbf{Input representation.} For a given patient-biomarker pair, we denote the ordered sequence of observed laboratory values as $\mathbf{x} = (x_1, \ldots, x_T)$, with corresponding measurement times $\mathbf{t} = (t_1, \ldots, t_T)$ and clinical states $\mathbf{s} = (s_1, \ldots, s_T)$, where each $s_i$ indicates the laboratory state relative to the population reference interval (low, normal, or high).

The model processes these inputs as a sequence of three token types: a context token, history tokens, and a query token. The context token $z_c$ encodes static patient covariates: sex $g$, age $a$, and laboratory test code $c$, as a sum of learned embeddings:
\begin{equation}
z_c = e_g(g) + e_a(a) + e_c(c).
\end{equation}

Each history token $z_i^{\text{hist}}$ represents one prior measurement and combines three components:
\begin{equation}
z_i^{\text{hist}} = f_v(\tilde{x}_i) + e_s(s_i) + f_\tau(\Delta t_i),
\end{equation}
where $f_v$ is a linear projection of the input value $\tilde{x}_i$, $e_s$ is a learned state embedding, and $f_\tau$ encodes the inter-measurement interval $\Delta t_i = t_i - t_{i-1}$ (Supplementary Methods). Alternative input encodings are summarized in Supplementary Table~X.

The query token, $z^{\text{query}}$, specifies the context at the time of prediction:
\begin{equation}
z^{\text{query}} = e_s(s_{T+1}) + f_h(t_{T+1} - t_T),
\end{equation}
where $s_{T+1}$ denotes the future health state and $t_{T+1} - t_T$ the prediction horizon. The full sequence is processed using a causally masked Transformer decoder to obtain the predictive distribution of the next observed laboratory value by $P(x_{T+1} \mid \mathbf{x}, \mathbf{t}, \mathbf{s}, g, a, c, s^*, t^*)$. Supplementary Table~3 summarizes the architectural and training configuration options, including temporal encoding, age representation, and output parameterization.

\textbf{Training objective.} Given the input sequence, the model is optimized to predict the conditional distribution of the next observed value $x_{T+1}$. We trained NORMA under two output parameterizations.

Under the Gaussian parameterization, the model jointly estimates a predicted mean $\mu$ and log-variance $\log \sigma^2$:
\begin{equation}
\mathcal{L} = \frac{1}{2}\left[\log \sigma^2 + \frac{(x_{T+1} - \mu)^2}{\sigma^2}\right].
\end{equation}

Under the quantile parameterization, the model directly predicts fixed quantiles $\tau \in \{0.025, 0.25, 0.50, 0.75, 0.975\}$ using a pinball loss without distributional assumptions:
\begin{equation}
\mathcal{L}_\tau = \tau \cdot \max(0, x_{T+1} - \hat{q}_\tau) + (1-\tau) \cdot \max(0, \hat{q}_\tau - x_{T+1})
\end{equation}
where $\hat{q}_\tau$ is the predicted $\tau$-th quantile. Both parameterizations share the same core architecture (embedding dimension 64, 4 attention heads, 8 Transformer layers) and were trained on the same development data.

\textbf{NORMA reference intervals.} We define the NORMA reference interval (NORMA-RI) as the 95\% prediction interval obtained by conditioning on a future normal laboratory state. Under the Gaussian parameterization, this corresponds to $[\mu - 1.96\sigma,\; \mu + 1.96\sigma]$; under the quantile parameterization, it corresponds to $[\hat{q}_{0.025},\; \hat{q}_{0.975}]$.

\textbf{Evaluation.} We first evaluated forecasting performance for next-step prediction. For point accuracy, we used the 50th percentile (median) prediction and compared it against three baselines: a person-specific historical mean model, last observation carried forward (``Last''), and an autoregressive integrated moving average model (ARIMA). Performance was quantified using mean absolute error (MAE), mean absolute percentage error (MAPE), and coefficient of determination (R$^2$).

\textbf{Sensitivity Analysis.} To assess how NORMA captures uncertainty across clinical contexts, we conducted one-at-a-time sensitivity sweeps over five input features: patient age, sex, number of prior measurements, prediction horizon, and within-sequence variability. For each of 30 laboratory tests, we constructed a synthetic baseline sequence: 10 measurements spaced 90 days apart, each set to the midpoint of the sex-specific population reference interval, for a 50-year-old male with a 30-day prediction horizon. We then varied one feature at a time while holding all others at baseline values. Age was swept from 20 to 80 years in 5-year increments; history length from 2 to 300 measurements; and prediction horizon from 7 days to 10 years. For within-sequence variability, we parameterized the standard deviation as a multiplier of one-tenth of the reference range width (0.0 to 3.0) and, at each noise level, drew 30 independent histories from $\mathcal{N}(\text{midpoint}, \sigma)$ to capture sampling variability. For each configuration, we recorded the predicted median (50th quantile), the 90\% prediction interval width, and the percent change in interval width relative to baseline, enabling direct comparison of sensitivity magnitude across biomarkers.

\subsection*{Clalit Health Services}

Clalit Health Services (CHS) is a nationwide health-system cohort from Israel with $>$20 years of longitudinal electronic healthcare data. We included adults aged 18--99 years with repeated outpatient laboratory testing between 2000 and 2024. For each analyte, eligibility required at least five outpatient measurements spaced at least 90 days apart prior to January 1, 2015 (baseline period) and at least one measurement between January 1, 2015 and January 1, 2016 (classification period). We excluded hospitalizations, emergency department visits, and urgent encounters during baseline ascertainment. The first measurement during the classification period served as the index value and defined the index date. We assessed eligibility independently per analyte, allowing individuals to contribute to multiple analyses.

We followed individuals for up to ten years from each analyte-specific index date. For those without events, we censored follow-up at the earliest of health plan disenrollment, death. If still enrolled in the health plan with no membership end date or date of death, we used the last available day in the dataset as the censoring date (January 1, 2024).

\textbf{Outcomes.} Primary outcomes included all-cause mortality and incident diagnoses of type 2 diabetes (T2D) and chronic kidney disease (CKD). We ascertained outcomes using ICD-9/10 codes supplemented by laboratory criteria. We defined T2D as HbA1c $\geq$6.5\% or fasting glucose $\geq$126 mg/dL on two occasions; CKD as estimated glomerular filtration rate $<$60 mL/min/1.73 m$^2$ on two measurements at least 90 days apart. We supplemented structured data with clinical text using predefined keyword lists and excluded negated and family-history mentions.

\textbf{NORMA configuration.} For the CHS validation, we applied NORMA with the default configuration: Time2Vec temporal encoding, ternary health-state indicators, raw-value laboratory inputs, raw age projected linearly, and the Gaussian output parameterization (Supplementary Table~3).

\subsection*{eICU Collaborative Research Database}

The eICU Collaborative Research Database (eICU-CRD v2.0) is a multi-center critical care dataset comprising over 200,000 ICU admissions from 208 hospitals across the United States. We included adults aged 18--99 years with repeated laboratory measurements during their ICU stay. We applied a baseline filter requiring at least five measurements. After filtering, 98,432 unique patients contributed over 20 million measurements across 30 analytes. Classification required all three methods (PopRI, PerRI, NORMA-RI) to have valid interval bounds, resulting in 54,418 patients with 2,931,704 classified index measurements. Baseline measurements (first 75\% of each patient-analyte sequence) were used to estimate PerRI and NORMA-RI; the remaining measurements served as index values for classification.

\textbf{Outcomes.} Primary outcomes included in-hospital mortality, acute kidney injury (AKI), sepsis, and prolonged ICU length of stay ($>$7 days). Outcomes were ascertained using ICD-9 diagnosis codes and structured clinical documentation.

\textbf{NORMA configuration.} For the eICU validation, we applied NORMA with the following configuration: log-delta-t temporal encoding with periodic components, ternary health-state indicators, within-sequence normalization of laboratory values, binned age embeddings (decade-wide bins), a dedicated context token for patient covariates, and the quantile output parameterization (Supplementary Table~3).

\subsection*{Statistical Analysis}

\textbf{Mortality association analysis.} We grouped index measurements into deciles of raw analyte values and estimated five-year mortality risk within each decile. We calculated 95\% confidence intervals using bootstrap resampling ($n = 1{,}000$).

\textbf{Abnormality classification.} We classified each index measurement as normal or abnormal under PopRI, PerRI, and NORMA-RI. To avoid incorporating pre-existing pathology into PerRI thresholds, we excluded patients whose PerRI mean fell outside the PopRI. Measurements classified as PopRI-abnormal were also classified as abnormal using PerRI and NORMA-RI.

\textbf{Clinical Risk Stratification and Prediction.} We evaluated whether abnormal classifications under each framework identified individuals at elevated risk of future events. For each analyte and framework (PopRI, PerRI, NORMA-RI), we defined a binary indicator of abnormality for values outside the corresponding interval.

For each analyte--outcome pair, we calculated positive predictive value (probability of the event given abnormal classification), sensitivity (probability of abnormal classification among those with the event), and specificity (probability of normal classification among those without the event). To isolate differences between interval paradigms, we restricted analyses to measurements within the PopRI, because values outside the PopRI necessarily fall outside PerRI and NORMA-RI.

We estimated associations using Cox proportional hazards models with a single binary abnormality indicator adjusted for age and sex. We split data into training (60 percent) and test (40 percent) sets stratified by event status and approximate event time in one-year bins. We evaluated performance at 1, 3, 5, and 10 years using time-dependent receiver operating characteristic curves and quantified uncertainty with 95 percent confidence intervals from 1,000 bootstrap resamples. We applied Benjamini--Hochberg false discovery rate correction to account for multiple comparisons across outcomes and analytes.

\subsection*{Data Availability}

This study utilizes two publicly available datasets for the development of NORMA: EHRSHOT and MIMIC-IV, both accessible to qualified researchers under their respective data-use agreements. The eICU Collaborative Research Database is also publicly available via PhysioNet. Clalit Health Services data are not publicly available.

\subsection*{Code Availability}

The full source code for the NORMA model, including data processing, training, and evaluation pipelines, is publicly available at \url{https://github.com/aashnapshah/NORMA}. The interactive web-based user interface for individualized laboratory interpretation is available at \url{https://norma-tpy0.onrender.com/}.


\clearpage
\input{figures}
\clearpage
\input{supp_figures}
\clearpage
\input{tables}

\end{document}
